---
title: "ESM244_lab_1_CAMILLE"
author: "Camille Herrera"
date: "January 17, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Attach packages
```{r}

library(tidyverse)
library(janitor)
library(MASS)
library(ggrepel)
library(RColorBrewer)
library(factoextra)
library(ggbiplot)
library(shinythemes)

```

If you want to override (set) a function manually throughout an entire workspace, you can assign the function name explicityly

```{r}

select <- dplyr::select

# or can just use "dplyr::select" every time you want to use it instead of a general overide

```

###1. Ordinal logistic regression for political party affiliation by ethnicity, age, sex

```{r}

pol_party <- read_csv("pol_party.csv") %>% 
  clean_names() %>% # janitor function for snake case
  select(participant_id:read_carefully)

pol_df <- pol_party %>% 
  select(birth_year, gender, ethnicity, political_party) %>% 
  filter(birth_year > 1900, political_party != "NA", political_party <= 7) %>% 
  mutate(age = 2016 - birth_year)

# For political party: 1 = very liberal, 4 = moderate, 7 = very conservative

# Coerce to factors (including ordered factor for the ranking of political party)

pol_df$political_party <- factor(pol_df$political_party, ordered = TRUE, levels = c("1","2","3","4","5","6","7")) # Check class and levels to ensure

# Then update the gender and ethnicity variables to factors (unordered)

pol_df$ethnicity <- factor(pol_df$ethnicity)
pol_df$gender <- factor(pol_df$gender) 

# Then we can run ordinal logistic regression, but we should do some visualization first.

```
 
Political Party: 1 is very liberal and 7 is very conservative
 
Basic data exploration and visualization
 
```{r}

counts_eth <- pol_df %>% 
  group_by(ethnicity, political_party) %>% 
  tally()

counts_g <- pol_df %>% 
  group_by(gender, political_party) %>% 
  tally()

#looking at counts_eth can see there are less than n=15 for most groups by Race/Political party except for white respondents

ggplot(pol_df, aes(x = political_party)) + #for a histogram only give it x values
  geom_histogram(aes(fill = ethnicity), stat = "count") +
  scale_fill_brewer(palette = "YlGnBu") +
  facet_wrap(~ethnicity, scales = "free")

# non-normally distributed responses, extremes on either end point with less in the middle
# very small sample sizes for NA, highest for white respondents
# using facet_wrap we can show a lot of info in a few easy ways

ggplot(pol_df, aes(x = political_party)) + #for a histogram only give it x values
  geom_histogram(aes(fill = gender), stat = "count") +
  scale_fill_brewer(palette = "YlGnBu") +
  facet_wrap(~ethnicity, scales = "free")

```

# THESE DATA ARE PRECARIOUS AND N IS TOO SMALL!!! FOR GROUPS

```{r}
### THIS IS NOT PUBLISHABLE DATA

#Ordinal logistic regression
pol_model <- polr(political_party ~ age + gender + ethnicity, data = pol_df)

summary(pol_model)

# coeffieicts are log odds and with respect to different factors
# if log odds co-efficient of 0 that means no effect

exp(coef(pol_model))

###

# Once we'd decided on a model, we can always make predictions uing it:

predictions <- predict(pol_model, type ="probs")

df <- data.frame(pol_df, predictions)
# view(df)
```
"genderMale 0.312954"" an increase in log odds relative to the reference level

as you compare male to female, all else the same....

"ethnicityNative American 1.129312" no level of meaning, only 6 data points out of thousands

"age 1.001290" no effect on data
baselevel ethnicity = Asian

###2. Principal components analysis (PCA)

Using WorldBank environmental indicators

SEE KEY to check out the data wrangling!

```{r}

wb_10 <- read_csv("wb_10.csv") %>% 
  select(-X1) %>% 
  column_to_rownames('code')

wb_pca <- prcomp(wb_10[3:8], scale. = TRUE) #for continuous variables

summary(wb_pca)

ggbiplot(wb_pca, groups = wb_10$region) +
  geom_text_repel(label = rownames(wb_10), size = 2, segment.size = 0.2) +
  theme_bw()

# factoextra package (built to help visualize outcomes of ordination methods)

# screeplot:

fviz_screeplot(wb_pca) # how much variance is explained by each of the principal components

# See contributions of variables to different components:

fviz_contrib(wb_pca, choice = "var", axes = 1, top = 6) #if had 100 variables may want to limit to top 10


fviz_contrib(wb_pca, choice = "var", axes = 2, top = 6) 

# Biplot (variables + points)
fviz_pca_biplot(wb_pca,
                col.var = "cos2",
                label = "var",
                repel = TRUE)


fviz_pca_biplot(wb_pca,
                col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             label = "var",
             repel = TRUE)

```
shows things that are simliar to each other (Chile, Ecudaor) versus different (USA)


